Climate-driven phytoplankton community variability in the Cariaco Basin, Venezuela

### 2025-07-15 agujournaltemplate.tex
Andrew's Feedback

\begin{abstract}
Phytoplankton communities play a crucial role in oceans' biogeochemical cycles. Climate drives changes in the physical environment that phytoplankton communities experience, thus impacting the productivity and diversity of marine ecosystems. Long-term time series can capture changes in climate and ecosystems, allowing us to untangle the effects on local communities. 
This study aims to describe and investigate the changes in the phytoplankton community of the Cariaco Basin, Venezuela, between 1995 and 2017. Water samples were taken during monthly cruises to determine microscopy cell counts and biogeochemical properties, which we integrated with data from the AMO and MEI v.2 climate indices and ERA5 climate model reanalysis data. 
Using clustering algorithms, we investigate the structure of the community data, revealing two clusters that describe a shift from high- to low-upwelling conditions in 2004, but also show a return to a community and conditions similar to those before the shift in 2014.
However,  community diversity did not follow the same pattern, showing a decrease between 1998 and 2009. To quantify the effect of climate and bottom-up drivers on phytoplankton community changes, we employed a gradient forest analysis and found that a combination of in situ upwelling-related variables (SST and nitrate concentration) and climate indices (AMO and MEI v.2) were the strongest predictors. Our study emphasizes the impact of long-term (decadal) climatic oscillations, such as the AMO, on the phytoplankton community of a tropical coastal ecosystem through modulation of the upwelling system and highlights the capability of ecosystems to recover from the effects of climate change.

@@

    % Phytoplankton is important
    Phytoplankton provide the foundation for marine food webs and play a crucial role in biogeochemical processes, including oxygen production and carbon export to the deep ocean \cite{falkowski_biogeochemical_1998}. Phytoplankton are a paraphyletic group of photo-autotrophic organisms exhibiting highly variable eco-physiological properties \cite{appeltans_magnitude_2012}. The short generation time and high diversity allow phytoplankton communities to respond relatively quickly to environmental changes, making them suitable indicators of ecosystem state \cite{alvarez-cobelas_what_1998, barton_anthropogenic_2016, di_cavalho_temporal_2023}.
    Coastal margins, where phytoplankton typically thrive, constitute only \qty{\sim 10}{\%} of the global ocean's surface but are responsible for an estimated \qty{40}{\%} of carbon sequestration to the seafloor \cite{yool_examination_2001, mullerkarger_importance_2005}.

    Data from remote sensing suggest a global increase in the frequency and spatial extent of coastal phytoplankton blooms over the last 20 years, but regional effects are highly variable \cite{dai_coastal_2023}.
    Climate change is predicted to have a lasting impact on phytoplankton communities, including shifts in community composition and diversity \cite{acevedo-trejos_glimpse_2014, boyd_biological_2016, henson_future_2021}. 
    Long-term time series of biological observations are crucial for understanding how marine ecosystems respond to short-term processes and long-term changes in the local physical environment and climate \cite{carstensen_need_2014, henson_observing_2016}. Such time series are rare, particularly in the tropics \cite{clarke_does_2017}, where community diversity is often highest \cite{brown_why_2014, righetti_global_2019} and faces greater threats of extinction compared to temperate regions \cite{finnegan_paleontological_2015}.

    A tropical coastal location that has been studied intensively in the last few decades is the Cariaco Basin, where a wide range of physical and biogeochemical parameters have been consistently collected from 1995 to 2017 \cite{muller-karger_scientific_2019}. 
    % Describe CARIACO location
    The Cariaco Basin, located off the coast of Venezuela in the southern Caribbean Sea, is a highly productive ecosystem within one of the largest anoxic basins in the world oceans \cite{edgcomb_accessing_2011}. The basin is a continental shelf depression comprising two sub-basins of \qty{\sim 1400}{m} depths, which are effectively separated from the Caribbean Sea by a sill with a mean depth of about \qty{100}{m}. The water below the sill is poorly ventilated. This underwater topography, in combination with high surface productivity and aerobic degradation of sinking organic matter, creates permanent anoxic conditions at depths below \qty{\sim 275}{m} \cite{thunell_organic_2000}. Due to these unique properties, the Cariaco Basin has been used as a natural laboratory since the 1950s, providing a rich sedimentary record for the study of past climate \cite{hughen1996nature} and serving as a testbed for understanding the coupling between pelagic production and carbon export to the deep ocean \cite{montes_vertical_2012}.

    % Describe general seasonal cycle with functional groups etc.
    The Cariaco Basin is characterized by a marked seasonal cycle of upwelling, which is driven by easterly trade winds, with the highest levels of primary production observed from December through May when the Inter-Tropical Convergence Zone (ITCZ) moves southward. A smaller wind-driven secondary upwelling event occurs often in July and August \cite{mullerkarger_annual_2001, astor_seasonal_2003}. During the upwelling period, the supply of deeper nutrient-rich waters supports high biological productivity, and the phytoplankton community is dominated by diatoms \cite{romero_seasonal_2009}, with dinoflagellates, coccolithophorids, and nanoplankton also contributing. In the low-wind, rainy season, the Cariaco Basin exhibits oligotrophic conditions, and the fraction of large phytoplankton cells is reduced. However, nano- and picophytoplankton cells are present in relatively high numbers throughout the year \cite{lorenzoni_characterization_2015}.

    % PARAGRAPH THREE - CHANGES IN CARIACO
    Since the inception of the time series in 1995, the ecosystem of the Cariaco Basin has undergone marked changes. \citeA{taylor_ecosystem_2012} analyzed the time series data up to 2010 and found significant decadal trends of increasing sea surface temperature (SST), coinciding with a decrease in upwelling and phytoplankton bloom intensities, which the authors attributed to global warming. During the initial phase of the time series (1996–2004), the Cariaco Basin underwent consistent and pronounced seasonal upwelling cycles, driven by strong trade winds during winter and spring (December – April) \cite{mullerkarger_annual_2001, astor_seasonal_2003}. After 2004, a significant reduction in the strength of trade winds during the upwelling period \cite{taylor_ecosystem_2012} caused a regime change in the physical and biogeochemical system. This strongly affected the phytoplankton community and the entire ecosystem. Microscopy cell counts indicated a drastic reduction of large phytoplankton, particularly diatoms, as confirmed, albeit not with the same high significance, by analysis of pigment data \cite{pinckney_phytoplankton_2015}. In general, a shift was observed in the phytoplankton community toward smaller cell sizes, along with increased abundances of coccolithophores, cryptophytes, and other phytoflagellates, groups commonly found in stratified water columns \cite{lorenzoni_characterization_2015}. An analysis of pigment data indicates a deepening of the chlorophyll maximum, an increase in diversity, and a reduction in the seasonal variability of community composition \cite{pinckney_phytoplankton_2015}.

    % PARAGRAPH FOUR - EXPLANATION OF CHANGES // OPEN QUESTIONS!!
    The shift in phytoplankton community composition associated with increased temperature and reduced upwelling intensities does not appear to be a permanent change in regime. Data from 2012 to 2017, the end phase of the CARIACO time series, indicates a return to strong upwelling intensities and to high numbers of diatoms and dinoflagellates \cite{muller-karger_scientific_2019}.
    The impact of large-scale climatic processes on the observed shifts in the phytoplankton community remains unclear. The Atlantic Multidecadal Oscillation (AMO) has emerged as an important factor influencing the climate system in the region and the ecosystems of the north Atlantic \cite{nye_ecosystem_2014}. The oscillations of the AMO index are strongly correlated with temperature indices in the Caribbean region \cite{stephenson_changes_2014}.
    The El Niño Southern Oscillation (ENSO) is teleconnected to the Caribbean Sea, given that during ENSO (indicated by a positive MEI v.2 index) sea surface temperatures increase and trade winds blow more to the north-northwest \cite{enfield_tropical_1997}. However, using the time series available only until 2010, \citeA{taylor_ecosystem_2012} found no significant correlation between MEI v.2 and in situ observations, and did not test for correlations with the AMO index.

@@

\section{Materials and Methods}
% Here is text on Materials and Methods.
%
\subsection{Study Site}
    %Talk about the Basin shortly, then about the time series
    The Cariaco Ocean Time Series Program was established to explore the relationship between ocean surface processes and the sinking flux of particulate carbon. The Cariaco Basin (Figure~\ref{fig:map}) is a large (approximately 160 km long and 70 km wide) and deep (approximately 1,400 m) basin, situated on the northeastern Venezuelan continental shelf, with a sill to the north, running almost parallel to the coastline at a mean depth of around \qty{100}{m}. This area is relatively productive ($\sim$ \qty{400}{g.C.m^{-2}.y^{-1}}), with primary production and vertical particulate organic matter fluxes exceeding those observed at BATS and HOT sites, resulting in the formation of anoxic conditions below \qty{\sim 275}{m} depth.

\begin{figure}
\noindent\includegraphics[width=\textwidth]{fig/Map_CAR.pdf}
\caption{Bathimetric map of the CARIACO Basin in the Caribbean Ocean, off the coast of Venezuela, indicating the CARIACO Mooring, the station of the Ocean Time-Series (located at \ang{10.50}N, \ang{64.67}W). A sill, running almost parallel to the coast at an average depth of \qty{100}{m}, from the Isla de Magarita and the Isla la Tortuga and beyond, encloses the Cariaco basin from the North.}
\label{fig:map}
\end{figure}

\subsection{In-Situ Data}
    Samples were collected as part of the Cariaco Ocean Time Series Program by means of monthly research cruises at a station located in the eastern Cariaco Basin (\ang{10.50}N, \ang{64.66}W) \cite{muller-karger_scientific_2019}.
    The CARIACO field program consisted of a monthly core oceanographic cruise with Niskin bottle and CTD deployments, along with other deployments and samples not utilized in this work. Field logistics and partial sample analysis were coordinated by the Fundación La Salle de Ciencias Naturales on Isla de Magarita, Venezuela. Details about the sampling procedures, analytical methods, data quality control, and inter-calibration procedures have already been described by \citeA{mullerkarger_annual_2001, thunell_particulate_2007, astor_yrene_m_handbook_2013, astor_interannual_2013, astor_sintesis_2014}.

    All samples were collected using a 12-bottle rosette system equipped with a CTD (SeaBird, model SBE 25) \cite{astor_yrene_m_handbook_2013}. Water samples for chlorophyll a and nutrient analysis were taken monthly during the first cast of each cruise. These samples were collected at eight standard depths within the upper 100 meters (1, 7, 15, 25, 35, 55, 75, and 100 m).

    Between late 1995 and early 1998, nutrient data was provided by William Senior (Universidad de Oriente, Cumana, Venezuela) and from mid-1997 to  present by Dr. K. Fanning (University of South Florida, St. Petersburg, USA). Samples from eight cruises were analyzed by both laboratories. Data for PO$_4$ and SiO$_4$ were in strong agreement (within ±5\%), whereas  NO$_\text{x}$ and NH$_4$ measurements showed discrepancies. \cite{taylor_ecosystem_2012}. We used a merged dataset for all nutrients, in which coincidental monthly measurements are averaged.

    For the identification of phytoplankton species by microscopy, \qty{500}{\milli\liter} seawater samples were collected at the same depths in HDPE bottles and preserved with a \qty{5}{\%} (final concentration) formalin solution neutralized with sodium tetraborate. Quantitative analysis was performed at the Universidad de Oriente, Venezuela, following the Utermöhl method \cite{hasle1978inverted}, using \qty{100}{\milli\liter} sedimentation chambers with a 48-hour settling period. Phytoplankton identification and enumeration by microscopy were restricted to microphytoplankton (cells $>$ \qty{20}{\micro \meter}) \cite{mutshinda_environmental_2013}.
    %from Mutshinda et al. 2013
    Quality control of the microscopy data was performed by assembling the taxonomic identifications made across all cruises, removing observations that were not identified at the species or genus level (with the exception of nanoflagellate cell counts) correcting spelling variations, and grouping synonyms using taxonomic information from the World Register of Marine Species (\url{marinespecies.org}). After processing the data, 482 phytoplankton species were observed over the course of the time series, which were grouped into 223 phytoplankton genera.

    Data collected with Niskin bottles, CTD, and phytoplankton taxonomy were retrieved from the Biological and Chemical Oceanography Data Management Office (BCO-DMO) \url{https://www.bco-dmo.org/project/2047}. 
    From the CARIACO Time Series data, we used only those variables that cover the entire length of the time series without large gaps (i.e. no more than three months) in coverage. In total, we used a complete set of 254 monthly data points, for which all variables were available.

\subsection{Climate data}
    %from Taylor et al. 2012
    Zonal wind stress (-u, easterly Trade Winds) across the southern Caribbean Sea drives nearshore upwelling of nutrient-rich waters via Ekman transport \cite{rueda-roa_southern_2013}. We extracted the -u wind component at \qty{10}{\meter} from the ECMWF Reanalysis v5 (ERA5) climate model and data reanalysis in the grid cell closest to the Cariaco time series sampling location (\ang{10.4}N, \ang{64.8}W, 31 km grid size), obtained from \url{https://cds.climate.copernicus.eu/datasets/reanalysis-era5-single-levels}. For simplicity, we refer to this variable as "wind speed" throughout the article. The data for precipitation and evaporation are also extracted from the ERA5 dataset for the aforementioned grid cell. Evaporation provides an estimate of the accumulated amount of water that has evaporated from Earth's surface, while precipitation is an estimate for total precipitation, comprising all water that falls on Earth's surface as rain or snow.

    The monthly AMO index data (10-year low-pass), which represents area-averaged, detrended low-pass filtered North Atlantic SST anomalies, were obtained from \url{https://climatedataguide.ucar.edu/climate-data/atlantic-multi-decadal-oscillation-amo}.

    The bi-monthly multivariate El Niño/Southern Oscillation (ENSO) indices (MEI v.2), defined as the time series of the leading combined Empirical Orthogonal Function (EOF) of five different variables (sea level pressure (SLP), sea surface temperature (SST), zonal and meridional components of surface wind, and outgoing long-wave radiation (OLR)) over the tropical Pacific basin (\ang{30}S-\ang{30}N and \ang{100}E-\ang{70}W), were obtained from \url{https://psl.noaa.gov/enso/mei/}. In our analysis, bi-monthly values are aligned with the tail month (December-January is treated as the value for January).

@@

    As the main indicator of diversity, we used genus richness, thereby reducing taxonomic resolution to create a more homogeneous metric that is robust to discrepancies in sampling effort \cite{ptacnik_diversity_2008, de2020higher}. Furthermore, we calculated the Shannon index $H' = -\sum_i p_i \log_{e} p_i$ and Pielou's evenness $J' = -\sum_i p_i ln( p_i )/ln(H')$, where $p_i$ represents the genera observed for each interpolated and integrated monthly sampling.

@@

    \subsubsection{Assessing Environmental Drivers of Phytoplankton Dynamics}
    To quantify the effects of bottom-up drivers in the Cariaco Basin on shifts in the phytoplankton community, we employed a Gradient Forest analysis. Gradient Forest is a community-level extension of random forest analysis that combines regression trees calculated for indicator variables (here: monthly observed genus counts) to assess turnover in indicator variables across predictor variables \cite{pitcher_example_2012, large_critical_2015, tam_comparing_2017}. Gradient Forest calculates a goodness-of-fit statistic for each predictor variable and indicator variable, then it averages these to determine the overall importance of each predictor. The regression tree split values also identify the ranges of predictors in which significant changes occur in the indicator variables, signaling a threshold for phytoplankton community turnover. Cumulative plots of the predictive importance of each predictor variable can reveal thresholds for ecosystem drivers, as steep increases in cumulative importance signal a threshold for community turnover in the phytoplankton community \cite{tam_comparing_2017}. This analysis was conducted using monthly measurements integrated over a depth of 100 m for observed genus counts, ensuring the inclusion of all available data. The drivers/predictor variables considered in the analysis were the AMO index, the MEI v.2 index, wind speed, evaporation, precipitation, \qty{21}{\celsius} isotherm, SST, salinity, and nutrient concentrations ($\mathrm{[NO]}_3$, $\mathrm{[PO]}_4$ and $\mathrm{[SiO]}_4$). 
    The environmental variables from ocean time series tend to be strongly correlated, which the Gradient Forest methodology accounts for by implementing a conditional permutation of the correlated predictors \cite{ellis_gradient_2012}. 
    We show a model run without any time lags in  supplementary material (Figure \ref{fig:sup:GFoutput_nolags}). To consider the temporal lag in effect, particularly of climate indices, we included time lags of up to 6 months for all climate variables, as well as 3-month lags for in-situ measurements in exploratory runs. We chose time lags that scored highest in the goodness-of-fit statistics for the final model run (Table \ref{tab:sup:GFlagtests} in Appendix).

@@

% Figure 2
Figure \ref{fig:zscore} shows the normalized yearly means of all environmental variables considered, the logarithmic transformation of chlorophyll \textit{a}, and phytoplankton cell counts per functional group. We included marks to highlight years with a significant number of missing monthly measurements ($=$ 2, 3 months and $\geq$ 4 months), since these make the annual means less reliable. Particularly toward the later years of the time series, the sampling became sparse. The patterns emerging from wind speed, evaporation, precipitation, SST, and \qty{21}{\celsius} isotherm depth clearly show the years of positive anomalies of wind (blue) and negative anomalies for SST (red) and upwelling (red, i.e., shallower depth of the isotherm). These patterns are reflected in the mode of the AMO, where a positive anomaly correlates with weak upwelling regimes (see also Figure \ref{fig:sup:correlation} in Appendix). The yearly mean nutrient concentrations have a maximum positive anomaly at the start of the time series but show differing inter-annual oscillations. Annual mean anomalies of nitrate and phosphate appear to follow upwelling anomalies but remain relatively low between 2002 and 2014, with some variability in mean phosphate concentration. Silicate does not follow a clear pattern but shows an elevated mean concentration between 2008 and 2013, which could be driven by increased surface runoff \cite{lorenzoni_characterization_2015} and reduced depletion due to diminished diatom abundances.

\begin{figure}
\noindent\includegraphics[width=\textwidth]{fig/Figure2_ZScores_v2.pdf}
\caption{Time series showing annual z-scores of selected indicator and driver variables. Red indicates negative annual anomalies and blue indicates positive annual anomalies. Deeper shades indicate larger anomalies. All data are available at monthly resolution. The years 1995 (n=2) and 2016 (n=5), 2017 (n=1) were removed due to lack of data, additionally we marked years with two or three and four or more missing values to highlight unreliable annual mean estimates. Diversity Indices are calculated based on microscopy counts at the Genus level.}
\label{fig:zscore}
\end{figure}

>>>>

\begin{abstract}
Phytoplankton communities play a crucial role in ocean biogeochemical cycles. Climate drives changes in the physical environment that phytoplankton communities experience, thus impacting the productivity and diversity of marine ecosystems. Long-term time series of XXX can capture changes in climate and ecosystems, allowing us to understand the effects of environmental change on local communities. This study aims to describe and investigate the changes in the phytoplankton community of the Cariaco Basin, Venezuela, between 1995 and 2017. Water samples were taken during monthly cruises to determine plankton community composition and local water column conditions, which we integrated climate indice and meteorological data. 
Using clustering algorithms, we investigate the structure of the plankton community data, revealing two clusters that described a shift from high- to low-upwelling conditions in 2004, but also show a return to a community and conditions similar to those before the shift in 2014. However,  community diversity did not follow the same pattern, showing a decrease between 1998 and 2009. To quantify the effect of climate and bottom-up drivers on phytoplankton community changes, we employed a gradient forest analysis and found that a combination of in situ upwelling-related variables (sea surface temperature and nitrate concentration) and climate indices (AMO and MEI v.2) were the strongest predictors. Our study emphasizes the impact of long-term (decadal) climatic oscillations, such as the AMO, on the phytoplankton community of a tropical coastal ecosystem through modulation of the upwelling system and highlights the capability of ecosystems to recover from the effects of climate change.

@@

    % Phytoplankton is important
    Phytoplankton provide the foundation for marine food webs and play crucial roles in biogeochemical processes, \cite{falkowski_biogeochemical_1998}.  Their short generation times and high assemblage diversity allow phytoplankton communities to respond relatively quickly to environmental changes, making them suitable indicators of ecosystem state and dynamics \cite{alvarez-cobelas_what_1998, barton_anthropogenic_2016, di_cavalho_temporal_2023}. Long-term time series of biological observations are crucial for understanding how phytoplankton communities respond to short-term processes and long-term changes in the local physical environment and climate \cite{carstensen_need_2014, henson_observing_2016}. Such time series are rare, particularly in the tropics \cite{clarke_does_2017}.

A tropical coastal location that has been studied intensively in the last few decades is the Cariaco Basin, where a wide range of physical and biogeochemical parameters were consistently collected from 1995 to 2017 \cite{muller-karger_scientific_2019}. The Cariaco Basin, located off the coast of Venezuela in the southern Caribbean Sea, is a highly productive ecosystem within one of the largest anoxic basins in the world oceans \cite{edgcomb_accessing_2011}. The Cariaco Basin is characterized by a marked seasonal cycle of upwelling, which is driven by easterly trade winds, with the highest levels of primary production observed from December through May when the Inter-Tropical Convergence Zone (ITCZ) moves southward. A smaller wind-driven secondary upwelling event occurs often in July and August \cite{mullerkarger_annual_2001, astor_seasonal_2003}. During the upwelling period, the supply of deeper nutrient-rich waters supports high biological productivity, and the phytoplankton community is dominated by diatoms \cite{romero_seasonal_2009}, with dinoflagellates, coccolithophorids, and nanoplankton also contributing. In the low-wind, rainy season, the Cariaco Basin exhibits oligotrophic conditions, and the fraction of large phytoplankton cells is reduced. However, nano- and picophytoplankton cells are present in relatively high numbers throughout the year \cite{lorenzoni_characterization_2015}.

    % PARAGRAPH THREE - CHANGES IN CARIACO
    Since the inception of the time series in 1995, the ecosystem of the Cariaco Basin has undergone marked changes. \citeA{taylor_ecosystem_2012} analyzed the time series data up to 2010 and found  decadal trends of increasing sea surface temperature (SST), coinciding with a decrease in upwelling and phytoplankton bloom intensities, which the authors attributed to global warming. During the initial phase of the time series (1996–2004), the Cariaco Basin underwent consistent and pronounced seasonal upwelling cycles, driven by strong trade winds during winter and spring (December – April) \cite{mullerkarger_annual_2001, astor_seasonal_2003}. After 2004, a significant reduction in the strength of trade winds during the upwelling period \cite{taylor_ecosystem_2012} caused a regime change in the physical and biogeochemical system. This strongly affected the phytoplankton community and the entire ecosystem. Microscopy cell counts indicated a drastic reduction of large phytoplankton, particularly diatoms, as confirmed, albeit not with the same high significance, by analysis of pigment data \cite{pinckney_phytoplankton_2015}. In general, a shift was observed in the phytoplankton community toward smaller cell sizes, along with increased abundances of coccolithophores, cryptophytes, and other phytoflagellates, groups commonly found in stratified water columns \cite{lorenzoni_characterization_2015}. An analysis of pigment data indicated a deepening of the chlorophyll maximum, an increase in species diversity, and a reduction in the seasonal variability of community composition \cite{pinckney_phytoplankton_2015}.

    % PARAGRAPH FOUR - EXPLANATION OF CHANGES // OPEN QUESTIONS!!
    However, the shift in phytoplankton community composition associated with increased temperature and reduced upwelling intensities in year-start-year-end did not appear to be a permanent change in regime. Data from 2012 to 2017, the end phase of the CARIACO time series, indicated a return to strong upwelling intensities and to high numbers of diatoms and dinoflagellates \cite{muller-karger_scientific_2019}. The impact of large-scale climatic processes on the observed shifts in the phytoplankton community remains unclear. The Atlantic Multidecadal Oscillation (AMO) has emerged as an important factor influencing the climate system in the region and the ecosystems of the north Atlantic \cite{nye_ecosystem_2014}. The oscillations of the AMO index are strongly correlated with temperature indices in the Caribbean region \cite{stephenson_changes_2014}. The El Niño Southern Oscillation (ENSO) is teleconnected to the Caribbean Sea, given that during ENSO (indicated by a positive MEI v.2 index) sea surface temperatures increase and trade winds blow more to the north-northwest \cite{enfield_tropical_1997}. However, using the time series available only until 2010, \citeA{taylor_ecosystem_2012} found no significant correlation between MEI v.2 and in situ observations, and did not test for correlations with the AMO index.

@@

\section{Materials and Methods}
% Here is text on Materials and Methods.
%
\subsection{Study Site}
    %Talk about the Basin shortly, then about the time series
    The Cariaco Ocean Time Series Program was established to explore the relationship between ocean surface processes and the sinking flux of particulate carbon. The Cariaco Basin (Figure~\ref{fig:map}) is a large (approximately 160 km long and 70 km wide) and deep (approximately 1,400 m) basin, situated on the northeastern Venezuelan continental shelf, with a sill to the north, running almost parallel to the coastline at a mean depth of around \qty{100}{m}. This area is relatively productive ($\sim$ \qty{400}{g.C.m^{-2}.y^{-1}}), with primary production and vertical particulate organic matter fluxes exceeding those observed at the BATS and HOT sites, resulting in the formation of anoxic conditions below \qty{\sim 275}{m} depth.

\begin{figure}
\noindent\includegraphics[width=\textwidth]{fig/Map_CAR.pdf}
\caption{Bathymetric map of the CARIACO Basin in the Caribbean Ocean. The station is located at \ang{10.50}N, \ang{64.67}W). A sill, running almost parallel to the coast at an average depth of \qty{100}{m}, from the Isla de Magarita and the Isla la Tortuga and beyond, encloses the Cariaco basin from the North.}
\label{fig:map}
\end{figure}

\subsection{In-situ biological and environmental data}
    <Add a topic sentence stating what in situ data are use for.> Samples were collected as part of the Cariaco Ocean Time Series Program on monthly research cruises at a station located in the eastern Cariaco Basin (\ang{10.50}N, \ang{64.66}W) \cite{muller-karger_scientific_2019}.
    The CARIACO field program consisted of a monthly core oceanographic cruise with Niskin bottle and CTD deployments, along with other deployments and samples not utilized in this work. Details about the sampling procedures, analytical methods, data quality control, and inter-calibration procedures have already been described by \citeA{mullerkarger_annual_2001, thunell_particulate_2007, astor_yrene_m_handbook_2013, astor_interannual_2013, astor_sintesis_2014}.

    All samples were collected using a 12-bottle rosette system equipped with a CTD (SeaBird, model SBE 25) \cite{astor_yrene_m_handbook_2013}. Water samples for chlorophyll-a and nutrient analysis were taken monthly during the first cast of each cruise. These samples were collected at eight standard depths within the upper 100 meters (1, 7, 15, 25, 35, 55, 75, and 100 m).

    Between late 1995 and early 1998, nutrient data, including X, Y, and Z, were provided by William Senior (Universidad de Oriente, Cumana, Venezuela) and from mid-1997 to  present by Dr. K. Fanning (University of South Florida, St. Petersburg, USA). Samples from eight cruises were analyzed by both laboratories. Data for PO$_4$ and SiO$_4$ were in strong agreement (within ±5\%), whereas  NO$_\text{x}$ and NH$_4$ measurements showed discrepancies. \cite{taylor_ecosystem_2012}. We used a merged dataset for all nutrients, in which coincidental monthly measurements are averaged.

    For the identification of phytoplankton species by microscopy, \qty{500}{\milli\liter} seawater samples were collected at the same depths in HDPE bottles and preserved with a \qty{5}{\%} (final concentration) formalin solution neutralized with sodium tetraborate. Species counts were performed at the Universidad de Oriente, Venezuela, following the Utermöhl method \cite{hasle1978inverted}, using \qty{100}{\milli\liter} sedimentation chambers with a 48-hour settling period. Phytoplankton identification and enumeration by microscopy were restricted to microphytoplankton (cells $>$ \qty{20}{\micro \meter}) \cite{mutshinda_environmental_2013}.
    %from Mutshinda et al. 2013
    Quality control of the microscopy data was performed by assembling the taxonomic identifications made across all cruises, removing observations that were not identified at the species or genus level (with the exception of nanoflagellate cell counts), correcting spelling variations, and grouping synonyms using taxonomic information from the World Register of Marine Species (\url{marinespecies.org}). 482 phytoplankton species were observed over the course of the time series, which were grouped into 223 phytoplankton genera.

    Data collected with Niskin bottles, CTD, and phytoplankton taxonomy were retrieved from the Biological and Chemical Oceanography Data Management Office (BCO-DMO) \url{https://www.bco-dmo.org/project/2047}. 
    From the CARIACO Time Series data, we used only those variables that cover the entire length of the time series without large gaps (no more than three months) in coverage. In total, we used a complete set of 254 monthly data points from YearX-YearY, for which all variables were available.

\subsection{Meteorological and climate data}
    %from Taylor et al. 2012
    In addition to oceanographic environmental data, we also examine meteorological and climatic data that are known to influence ocean conditions and marine ecosystems. Zonal wind stress (u, easterly Trade Winds) across the southern Caribbean Sea drives nearshore upwelling of nutrient-rich waters  \cite{rueda-roa_southern_2013}. We extracted the u wind component at \qty{10}{\meter} from the ECMWF Reanalysis v5 (ERA5) climate model and data reanalysis in the grid cell closest to the Cariaco time series sampling location (\ang{10.4}N, \ang{64.8}W, 31 km grid size), obtained from \url{https://cds.climate.copernicus.eu/datasets/reanalysis-era5-single-levels}. For simplicity, we refer to this variable as "wind speed" throughout the article. The data for precipitation and evaporation are also extracted from the ERA5 dataset for the aforementioned grid cell. Evaporation provides an estimate of the accumulated amount of water that has evaporated from Earth's surface, while precipitation is an estimate for total precipitation, comprising all water that falls on Earth's surface as rain or snow.

    In addition the wind data, we examined two climatic indices, each of which provide information on the basin-scale climatic conditions and consequently the broad-scale forcing of marine ecosystems. The monthly Atlantic Multidecadal Oscillation (AMO) index data (10-year low-pass), which represent area-averaged, detrended low-pass filtered North Atlantic SST anomalies over what area, were obtained from \url{https://climatedataguide.ucar.edu/climate-data/atlantic-multi-decadal-oscillation-amo}.

    The bi-monthly multivariate El Niño/Southern Oscillation (ENSO) indices (MEI v.2), defined as the time series of the leading combined Empirical Orthogonal Function (EOF) of five different variables (sea level pressure, SLP); sea surface temperature, SST; zonal and meridional components of surface wind; and outgoing long-wave radiation, OLR)) over the tropical Pacific basin (\ang{30}S-\ang{30}N and \ang{100}E-\ang{70}W), were obtained from \url{https://psl.noaa.gov/enso/mei/}. In our analysis, bi-monthly values were aligned with the tail month (December-January is treated as the value for January).

@@

    We calculated three metrics of phytoplankton community diversity. First, we calculated genus richness, which reduces taxonomic resolution cut creates a more homogeneous metric that is robust to discrepancies in sampling effort \cite{ptacnik_diversity_2008, de2020higher}. Second and third, we calculated the Shannon index $H' = -\sum_i p_i \log_{e} p_i$ and Pielou's evenness $J' = -\sum_i p_i ln( p_i )/ln(H')$, respectively, where $p_i$ represents the genera observed for each interpolated and integrated monthly sampling.

@@

    \subsubsection{Assessing Environmental Drivers of Phytoplankton Dynamics}
    To quantify the effects of bottom-up drivers in the Cariaco Basin on shifts in the phytoplankton community, we employed a Gradient Forest analysis. Gradient Forest is a community-level extension of random forest analysis that combines regression trees calculated for indicator variables (here: monthly observed genus counts) to assess turnover in indicator variables across predictor variables \cite{pitcher_example_2012, large_critical_2015, tam_comparing_2017}. Gradient Forest calculates a goodness-of-fit statistic for each predictor variable and indicator variable, then it averages these to determine the overall importance of each predictor. The regression tree split values also identify the ranges of predictors in which significant changes occur in the indicator variables, signaling a threshold for phytoplankton community turnover. Cumulative plots of the predictive importance of each predictor variable can reveal thresholds for ecosystem drivers, as steep increases in cumulative importance signal a threshold for community turnover in the phytoplankton community \cite{tam_comparing_2017}. This analysis was conducted using monthly measurements integrated over a depth of 100 m for observed genus counts, ensuring the inclusion of all available data. The drivers/predictor variables considered in the analysis were the AMO and MEI v.2 indices, wind speed, evaporation, precipitation, \qty{21}{\celsius} isotherm, SST, salinity, and nutrient concentrations ($\mathrm{[NO]}_3$, $\mathrm{[PO]}_4$ and $\mathrm{[SiO]}_4$). 
    The environmental variables from ocean time series tend to be strongly correlated, which the Gradient Forest methodology accounts for by implementing a conditional permutation of the correlated predictors \cite{ellis_gradient_2012}. 
    We show a model run without any time lags in  supplementary material (Figure \ref{fig:sup:GFoutput_nolags}). To consider the temporal lag in effect, particularly of climate indices, we included time lags of up to 6 months for all climate variables, as well as 3-month lags for in-situ measurements in exploratory runs. We chose time lags that scored highest in the goodness-of-fit statistics for the final model run (Table \ref{tab:sup:GFlagtests} in Appendix).

@@

Create subsections in results. This one could be "Interannual changes in the environment and ecosystem"
% Figure 2
Coordinated, interannual changes occur in normalized yearly means of all environmental variables considered, the logarithmic transformation of chlorophyll \textit{a}, and phytoplankton cell counts per functional group (Figure \ref{fig:zscore}). The patterns emerging from wind speed, evaporation, precipitation, SST, and \qty{21}{\celsius} isotherm depth clearly show the years of positive anomalies of wind (blue) and negative anomalies for SST (red) and upwelling (red, i.e., shallower depth of the isotherm). These patterns are reflected in the mode of the AMO, where a positive anomaly correlates with weak upwelling regimes (see also Figure \ref{fig:sup:correlation} in Appendix). The yearly mean nutrient concentrations have a maximum positive anomaly at the start of the time series but show differing inter-annual oscillations. Annual mean anomalies of nitrate and phosphate appear to follow upwelling anomalies but remain relatively low between 2002 and 2014, with some variability in mean phosphate concentration. Silicate does not follow a clear pattern but shows an elevated mean concentration between 2008 and 2013, which could be driven by increased surface runoff \cite{lorenzoni_characterization_2015} and reduced depletion due to reduced abundances of diatoms.  <I think the most obvious pattern in Fig. 2 has not yet been discussed: Chl and phyto are high in 1996-2004, then low in 2004-2013, followed by a switch in 2014-2015. Diversity is high when chl/phyto are low. I think you'll want to describe these patterns as they are interesting.>

\begin{figure}
\noindent\includegraphics[width=\textwidth]{fig/Figure2_ZScores_v2.pdf}
\caption{Time series showing annual z-scores of selected environmental, climatological, and ecological variables. Red indicates negative annual anomalies and blue indicates positive annual anomalies. Deeper shades indicate larger anomalies. The years 1995 (n=2) and 2016 (n=5), 2017 (n=1) were removed due to insufficient monthly data to generate robust annual anomalies. Additionally we marked years with two or three and four or more missing monthly values. Diversity indices were calculated based on microscopy counts at the genus level.}
\label{fig:zscore}
\end{figure}
